The design and implementation of standards-based Grid single sign-on using federated identity

### 2009-04-30 doc/sec/papers/GRID09/paper.tex
Editorial corrections, some new references; still not finished (~30% left)

\begin{abstract}
Security infrastructure is one of the most challenging tasks in the developement, integration and
deployment of Grid middlewares. Even though the Grid community addresses the security issue 
through public key infrastructures (PKI) to support mutual authentication using X.509 certificates,
maintaining the X.509 credential is not that easy for non-IT specialists, and is proved to be an 
obstacle for the more widely deployment of Grid technology. Considering the identity federation 
technology is a emergingly widely used technology which can facilitate cross-domain single sign-on,
without requiring the users to maintain the X.509 credentials rather than their own institutional
accounts, we believe utilizing identity federation for Grid middlwares is a promising direction 
for Grid technology to be more widely used.
This paper describes a single sign-on infrastructure developed as part of the NorduGrid/ARC grid 
middleware. It adopts the identity federation standard (SAML), as well as Web Service standand. 
It focus on single sign-on solution on the middleware level for users to access Grids by only 
using their frequently used accounts, without being bothered to maintain X.509 credentials. 
Unlike other research which utilize identity federation on the Grid portal level, solution on
the middleware level can provide more flexibility.
Also, the performance of single sign-on solution is measured. We identify the performance limitation 
of security-related services inside this solution, and analyze the ways to avoid the limitation.
To our knowledge, the work described in this paper is the first evaluated implementation that utilizes 
identity federation for Grid usage on the middlware level.
\end{abstract}

%-------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
Grid infrastructures facilitate the interoperability between wide-scale, cross-domain heterogeneous 
resources, as well as the accessibility provided to users. In terms of security issue, the Grid security 
infrastruture should provide accessibility as much as possible without loosing the security benifits.

The current Grid community uses Grid Security Infrastructure (GSI)~\cite{IFoster98} as the de facto 
standard for authentication and transport level security, which builds on the Public Key Infrastructure
(PKI) to support authentication using X.509 certificate. Mutual authentication is required 
by GSI based Grid deployment. Mutual authentication means that Grid users have to possess X.509 
credentials. To obtain a X.509 credential, a user needs to contact the Registration Authority (RA)
which will check the user's information and approve his certificate request; afterwards,
the Certificate Authority (CA) can issue the certificate according to the approval from 
the RA. The whole process is arduous and may take some time.
Meanwhile, to maintain a X.509 certificate is also not so easy for the non-IT community to deal 
with, since users have to periodically create proxy certificates by using Grid-specific command line 
utilities, such as grid-proxy-init or voms-proxy-init. Also maintaing the CAs or RAs is not a  
easy task, especially when the community of Grid users is getting much wider.

On the other hand, users normally could already have had some frequently-used campus or institutional 
accounts/credentials such as username/password. We believe that enabling Grid users to use their institutional 
credentials rather than the X.509 certificate to access Grids is a promising way 
to make the Grids more easily accessible, as well as to extend the user community of Grids
to a much larger scale.

To enable users to utilize the institutional credentials to access Grid, some issues needs to be 
addressed. Firstly, the Grid middleware needs to be improved to support the authentication based on
institutional credentials instead of X.509 certificates; Secondly, in order to interoperate with
those Grid infrastructures which require X.509 certificate based mutual authentication, an approach
should be provided for obtaining X.509 certificate based on the institutional credentials; Thirdly,
the single sign-on characteristic should be provided so that users can authenticate once and then transverse 
from Grid resources to Grid resources without being prompted to authenticate again at each of the resources.

The approach in the new version of ARC grid middleware is to utilize the identity federation 
standards, specifically, the Security Assertion Markup Language (SAML) specification, and Web 
Service standards, to facilitate the Grid single sign-on.

The rest of this paper is organized as follows: Section \ref{sec:arcmiddleware} presents the 
ARC Grid middleware, especially the architecture of the new version of ARC. Section 
\ref{sec:intergrationSAML2SSO} describes the solution of using federated identity for 
Grid authentication by using SAML2SSO profile; Setion \ref{sec:fedidtoX509} describes
how to obtain a X.509 credential based on federated identity, and how to use it for 
accessing multiple Grid resources; Section \ref{sec:perfeval} is performance evaluation;
Section \ref{sec:relatedwork} describes the related work; Section \ref{sec:conclusion} 
contains conclusions and future work outlook.

@@

ARC (Advanced Resource Connector) is an open source Grid middleware solution released 
under Apache license. ARC middleware aims at developing self-organized, 
fault-tolerant, non-intrusive, easy-manageable Grid middleware~\cite{MEllert07}. 
The current production version of ARC provides Grid services for job submission and 
management, resource characterization, resource aggregation and discovery, basic data 
management, integration of Grid security solution, and so on. This release has been 
deployed and used in production environment, and is one of the widely deployed Grid 
middlewares in Europe.

The development version of ARC components is developed by the KnowARC project~\cite{KnowARClink}, 
based on the functionality of the current production ARC middleware. It aims at 
implementing a Web Service oriented Grid middleware which will provide higher levels of 
resource and user abstraction through well-defined Web Service interface~\cite{KnowARCDesignlink} 
in order to provide interoperability with other Web Service oriented Grid middlewares, as well as 
other Web Service compatible applications.

As the key part of the implementation of new ARC middleware, there is a lightweight Web 
Service container called Hosting Environment Daemon (HED) which provides hosting environment 
for various services at application level, as well as a number of modules to support flexible, 
interoperable, and efficient communication mechanism for building SOAP based Web Services. 
The design of the HED is built around the idea of flexibility and modularity,
such that the application service developers can simply concentrate on the application 
level Web Service implementation by only using the core minimum amount of components. It 
also simplifies work on the middleware level, e.g. making possible to implement another 
communication protocol or authentication mechanism. A system administrator can easily 
configure and deploy the middleware and application for different kinds of requirements 
without having to know much about the implementation.

The architecture of the HED is illustrated in Figure \ref{fig:HED}. In general, there 
are few components called Message Chain Component (MCC) which are in charge of implementing 
different protocol levels. For instance, as shown in the example message flow, the HTTP MCC 
will process stream from the TLS MCC to parse the HTTP message and pass its body to the SOAP 
MCC, and also process the SOAP response from the SOAP MCC to generate the HTTP message 
for the TLS MCC.

Dotted line in Figure \ref{fig:HED} shows an alternative path for the information 
to propagate among MCCs. A service administrator can configure the MCCs according to the 
interoperability requirements with the other part. For instance, the configuration marked 
with the dotted line is compatible with the WSE (Web Services Enhancement for .NET) SOAP 
message mechanism (see WSE’s SoapSender and SoapReceiver)~\cite{WSElink}. Another configuration 
could be SOAP over HTTPG (HTTP over GSI) which is needed to interoperate with services 
like the Storage Resource Manager (SRM)~\cite{A.Shoshan03} service. This shows the 
flexibility of HED in terms of protocols support.

@@

HED contains a flexible security framework for implementing and enforcing security 
related functionality, such as authentication and authorization. 
Each security related functionality can be implemented as a pluggable and configurable 
component (plug-ins) called SecHandler. Each MCC or Service is usually configured with 
two queues of SecHandler -- one for the incoming message and one for the outgoing message 
respectively. In Figure \ref{fig:HED}, the ``AuthZ'' and ``AuthN'' sub-modules inside 
MCCs and Services are examples of SecHandler.

%-------------------------------------------------------------------------
\section{Grid authentication using federated identity: Integrating SAML2SSO profile}
\label{sec:intergrationSAML2SSO}
Identity federation is an emerging technology which enables the identity information to be 
trustily transfered across autonomous security domains. By using identity federation, users 
from one domain can access another domain without the need for directly trust relationship 
between users and accessed domain, i.e., users can use accounts from their host domains to 
access external domains, and accessed domain does not need to maintain accounts for those 
external users. Identity federation doesn't enforce specific authentication mechanisms, so 
that various types of authentication can be supported between user and its host domain, and 
user is able to use his frequently used account/credential to accomplish authentication. 
Considering the Grid use cases, we see that utilizing the existing account to access Grid 
without being bothered to applying and manipulating a X.509 credential should be attractive 
for Grid users.

Shibboleth is the one of the several implementations of the identity federation, which is 
based on the OASIS Security Assertion Markup Language (SAML) 2.0 specifications. In terms 
of authentication, the Shibboleth implements SAML2.0 web browser SSO profile, which  defines 
two functional components, an Identity Provider and a Service Provider. The Identity Provider 
(IdP) is responsible for creating, maintaining, and managing user identity, while the Service 
Provider (SP) is responsible for controlling access to services and resources by analyzing 
the SAML assertions produced and issued by the IdP upon request (request from client application 
to the service/resource that is protected by SP).

In the implementation of ARC middleware, SAML2.0 web browser SSO profile is supported by 
implementing the functionality of service provider and web browser's user agent, and utilizing 
Shibboleth IdP implementation for the functionality of identity provider. The user agent 
functionality is for mimicing the behavior of web browser's user agent for authentication, 
such as the HTTP redirection and the HTTP cookie processing. The implementation of the user 
agent is based on the the HTTPS client interface of ARC, since the client interface of ARC 
can support different protocols which are incarnated by different MCCs.

@@

Figure \ref{fig:SAML2SSOUML} shows how is SAML2.0 SSO profile integrated into the SOAP 
invocation between ARC WS Client and Web Service. Step 1 to step 5 in Figure \ref{fig:SAML2SSOUML} 
is similar to those steps depicted in Web Browser SSO profile[ref], with the differences that
we specify the way about how does the service provider determine which identity provider to use (identity
provider discovery), by enabling the client to send the IdP name to service provider; as well 
as specify the way about how does the client/user agent authenticate against service provider, 
by implementing the form-based HTTP authentication which uses username and password for authentication between 
user agent and service provider (rather than X.509 client authentication) and is compatible 
with the Username/Password login handler of Shibboleth IdP.

In step 6, the service provider (SP Service) verifies and checks the SAML response, decrypts and stores 
the SAML assertion into session/connection context. That assertion includes the AuthnStatement and 
AttributeStatement elements. From step 8 to step 10, the WS client launches the SOAP invocation via 
the same connection as the one which is used by user agent to contact SP Service, then Grid/Web Service 
checks the AuthnStatement stored in the session context to see if the AuthnStatement is valid or expired. 
If valid, service handles the SOAP request and returns the SOAP response to the WS client. Note
the ``connection'' and ``session'' are interchangeably used in this paper, with the same meaning for 
TCP connection.

Note that in the current implementation the service requires WS client to reuse the same 
TCP connection as the one used by user agent in step 5, in order to guarantee that the validity of 
the SAML2SSO result is applicable to the SOAP invocation process. On the service side, 
the sesion reusing is accomplish by deloying the SP service and other functional Grid/Web Service(s) 
together in the same service container.

There are a few benifits about integrating SAML2.0 SSO profile for Grid/Web service authentication. 
Firstly, the client certificate authentication can be switched off, so that users do not need 
to apply and maintain a X.509 credential. Secondly, some implementation of identity provider, such as 
Shibboleth IdP, can cache the authentication result through session management once the user
has succeeded to authenticate, and for a short period this authentication result is valid, so that 
the next time the user authenticates against IdP, the user agent can just feed IdP with the last 
successfully authenticated session's identity rather than feed IdP with username and password again. So 
the user (or the client on behalf of that user) can travel across multiple security domains with 
only providing his name and password once, which is also the characteristic of single sign-on.
Lastly, even though there are a few implementations of identity providers, most of them are 
standard-compliant, so the solution implemented in ARC middleware can interoperate with other 
identity provider implementations with minimum change.

%-------------------------------------------------------------------------
\section{Bridging federated identity and X.509 credential}
\label{sec:fedidtoX509}
Many widely used Grid middlewares are based on GSI which requires mutual X.509 authentication. 
Meanwhile, Web Service based Grid applications mostly do require client certificate authentication.  
To bridge the gap beween federated identity and X.509 credential, a short lived credential 
service (SLCS) is implemented through which a user can get a short-lived X.509 certificate only by
providing his username/password and authenticating through the SAML2SSO profile.

Also, in order to provide single sign-on capability for services that act on behalf of user to 
invoke other services, a X.509 credential delegation service is implemented, which is based on 
SOAP communication rather than the proprietary GSI communication which is used in the delegation
services such as MyProxy.

\subsection{Short-lived credential service}
\label{sec:slcs}
The sequence diagram of SLCS process is the same as the diagram shown in 
Figure \ref{fig:SAML2SSOUML}, with SOAP request (the message of step 8 of 
Figure \ref{fig:SAML2SSOUML}) including X.509 request 
and SOAP response (the message of step 10 of Figure \ref{fig:SAML2SSOUML}) 
including an signed X.509 certificate and the CA certificate that is used to 
sign this X.509 certificate. 
In detail, the SLCS client firstly accomplishes the SAML2SSO profile, then sends 
SOAP request to the SLCS service with the X.509 request embedded. The SLCS service makes
access control decision according to the SAML assertion stored in session context, and issues 
a certificate (with 12 hours lifetime) with the SAML attributes as the X.509 certificate 
extension. The SLCS client then gets the SOAP response with the X.509 certificate embedded.

Since the lifetime of the short lived credential has a short lifetime, it is permissible to protect 
the private key by using the local file system permissions instead of encrypting it by 
using a pass-phrase. Therefore, no manual interactivity is required to uncrypted the private
key. Hence single sign-on is provided, since the user only needs to input his password when 
he starts SAML2SSO profile.

How to compose a distinguished name (DN) for the certificate is a critical issue for 
the SLCS service. Since the ``eduPersonPrincipalName'' (see eduPerson schema~\cite{eduSchemalink})
is identical to the user that accomplishes SAML2SSO profile. We pick up ``eduPersonPrincipalName'' 
from SAML attributes that returned from Shibboleth IdP, and use it to compose the DN. A 
typical example of the eduPersonPrincipalName 
could be alice@example.org. In such case composed DN is for example\\``/O=knowarc/OU=example.org/CN=alice''.

By using the SLCS service, the user can easily access the Grid services that requires 
X.509 credential on client side, at anywhere, simply by running the SLCS client command 
and providing his username and password.

@@

The sequence diagram is shown in Figure \ref{fig:DelegUML}. Firstly, the client needs to delegate
the proxy certificate or short-lived certificate to a delegation service through step 1 to step 7.
Then this client invokes the peer Web Service by appending the delegation information (URL of delegation
service and the delegation ID corresponding to the above delegation) into the SOAP body. The Web 
Service then sends a X.509 request and the delegation ID to the delegation service 
which is specified by client side, and gets back a delegated certificate which can be composed together
with the private key for a proxy certificate. The Web Service could use this proxy certificate 
to represent the user to invoke another Web Service, which either could cause one more process of 
delegation (step 1 to step 13) if another delegation service is specified, or could directly start
from step 8 if the same delegation service is specified as the one in the former delegation process.

The client only needs to delegate the certificate to delegation service one time per session, it does
not need to do delegation for subsequent SOAP invokations after the first one.

The functionality of both client and service sides are implemented as pluggable handlers which can be 
configured into the client and service configuration, so that the service or client implementation itself
does not need to be changed.

We suppose the short-lived certificate is used by client, so that the SLCS service together with
the delegation service can provide single sign-on from user's federated identity, while keeping the 
interoperability to those Grid/Web services that requires mutual authentication using certificate.

%-------------------------------------------------------------------------
\section{Performace Evaluation}
\label{sec:perfeval}
The following sections evaluate the performance of the HED versus Axis2/C, as 
well as the performance about the SAML2SSO profile, short-lived credential
service, and delegation service.

\subsection{Performance comparison between HED and Axis2/C}
\label{sec:perhedandaxis}
Since the implementation of HED includes an alternative SOAP implementation which is the base of 
the security related web servies described in this paper, it is useful to compare the performance 
of HED and other SOAP implementations. Although there are many SOAP implementations such as 
AxisJava, gSOAP, XSOAP4, etc., the goal of this paper is not about complete comparison between
HED and all the other SOAP implementations, so we use the Axis2/C (v1.5.0) to demostrate the performance
HED.

For the service side, we implement two simple ``echoString'' services in HED and Axis2/C. The 
``echoString'' provides a simple SOAP portType which will output the same value as it gets from
input.
For client side, we develope a client by using the client API of ARC, to invoke the ``echoString''
services from Axis2/C and HED. The service invocation is just sending a short string to service, and 
getting back the same string. One test/experiment consists of a set of clients which run simultaneously. 
Each client runs for a fixed period of time, and invokes the service as many times as it can during 
the test period. Multiple clients are launched by creating multiple threads in one test.
Two values are measured: average response time and throughput. Average response time is computed
by counting the average value of the response time for each invocation. Throughput is computed
by counting the number of invocation during each second.

The two ``echoString'' services run in a test machine equipped with 3GHz Intel Pentium dual-core 
CPU and 2GB memory, Red Hat Enterprise Linux 5. The client runs in another machine equipped 
with 2.0GHz AMD Athlon CPU and 1GB memory, Red Hat Enterprise Linux 5. The Axis2/C service is run using 
the httpd module on Apache Http Server version 2.2.11 instance. We keep all the configurations 
default in Apache Http Server.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{TCP_thread_all.pdf}
\caption{Performance of HED and Axis2C echoString services without TLS: performance
changes with client numbers; session non-reused }
\label{fig:TCP_thread_all}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{TLS_thread_all.pdf}
\caption{Performance of HED and Axis2C echoString services with TLS: performance
changes with client numbers; session non-reused}
\label{fig:TLS_thread_all}
\end{figure}

Figure \ref{fig:HED2Axis_thread10} is the results of performance comparison with ten 
clients being launched.
Two types of experiments are shown: session reused, session non-reused.
``Session reused'' means all of the SOAP invocations on each client share the same session 
(tcp connection); while ``Session non-reused'' means each invocation launches a new session.

>>>>

\begin{abstract}
Security infrastructure is one of the most challenging tasks in the development, integration and
deployment of Grid middlewares. Even though the Grid community addresses the security issue 
through public key infrastructures (PKI) to support mutual authentication using X.509 certificates,
maintaining X.509 credentials is not that easy for non-IT-experts, and has proved to be an 
obstacle for a more wide deployment of Grid technologies. The identity federation 
is an increasingly popular technology that can facilitate cross-domain single sign-on
without requiring the users to maintain any credentials additional to their own institutional
accounts. We believe that utilizing identity federation for Grid middlewares is a promising path 
for the Grid technology to get more widely used.
This paper describes a single sign-on infrastructure developed as a part of the NorduGrid ARC Grid 
middleware. It adopts the identity federation standard (SAML), as well as Web Service approach. 
It focuses on a single sign-on solution at the middleware level for users to access Grids by only 
using their frequently used accounts, without being bothered to maintain X.509 credentials. 
Unlike other research which utilize identity federation on the Grid portal level, solution on
the middleware level can provide more flexibility.
In addition, the performance of single sign-on solution is measured. We identify performance limitations 
of security-related services inside this solution, and analyse the ways to avoid the limitations.
To our knowledge, the work presented in this paper is the first evaluated implementation that utilizes 
identity federation for Grid usage on the middleware level.
\end{abstract}

%-------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
Grid infrastructures facilitate interoperability between wide-scale, cross-domain heterogeneous 
resources, as well as the user access to these resources. In terms of security issues, a Grid security 
infrastructure should provide as simple accessibility as possible without loosing the security benefits.

Many current Grid communities predominantly use the Grid Security Infrastructure (GSI)~\cite{IFoster98} as the \textit{de facto} 
standard for authentication and transport level security. It builds on the Public Key Infrastructure
(PKI) to support authentication using X.509 certificates. GSI-based Grid deployment requires mutual authentication.
Mutual authentication means that Grid users have to possess X.509 
credentials. To obtain an X.509 credential, a user needs to contact a designated Registration Authority (RA)
which will check the user's personal data and eventually approve her certificate request; afterwards,
the respective Certificate Authority (CA) can issue the certificate following the approval from 
the RA. The whole process is arduous and may take quite some time.
Meanwhile, to maintain an X.509 certificate is also not so easy for the non-IT-skilled community to deal 
with, since users have to periodically create proxy certificates by using Grid-specific command line 
utilities, such as \texttt{grid-proxy-init} or \texttt{voms-proxy-init}. Also, operating CAs or RAs is not an  
easy task, especially when the community of Grid users is getting much larger.

On the other hand, users normally would already have some frequently-used campus or company 
accounts/credentials such as username/password. We believe that enabling Grid users to use their institutional 
credentials rather than the X.509 certificate to access Grids is a promising way 
to make the Grids more easily accessible, as well as to extend the user community of Grids
to a much larger scale.

To enable users to utilize the institutional credentials to access Grid, some issues need to be 
addressed. Firstly, the Grid middleware needs to be extended to support the authentication based on
institutional credentials instead of X.509 certificates; secondly, in order to interoperate with
those Grid infrastructures which require X.509 certificate-based mutual authentication, an approach
should be provided for obtaining X.509 certificate based on the institutional credentials; thirdly,
the single sign-on characteristic should be provided so that users can authenticate once and then traverse 
from Grid resources to Grid resources without being prompted to authenticate again at each resource.

The new approach in the ARC~\cite{MEllert07} Grid middleware is to utilize the identity federation 
standards, specifically, the Security Assertion Mark-up Language (SAML) specification, and Web 
Service standards, to facilitate the Grid single sign-on.

The rest of this paper is organized as follows: Section~\ref{sec:arcmiddleware} presents the 
ARC Grid middleware, and in particular its new architecture. Section~\ref{sec:intergrationSAML2SSO} 
describes the solution of using federated identity for 
Grid authentication by using \texttt{SAML2SSO} profile; Section~\ref{sec:fedidtoX509} describes
how to obtain X.509 credentials based on federated identity, and how to use such for 
accessing multiple Grid resources; Section~\ref{sec:perfeval} is dedicated to performance evaluation;
Section~\ref{sec:relatedwork} describes the related work; Section~\ref{sec:conclusion} 
contains conclusions and future work outlook.

@@

ARC (Advanced Resource Connector)~\cite{MEllert07} is an open source Grid middleware solution released 
under Apache license. ARC middleware aims at developing self-organized, 
fault-tolerant, non-intrusive, easy-manageable Grid middleware. 
The current production version of ARC provides Grid services for job submission and 
management, resource characterization, resource aggregation and discovery, basic data 
management, integration of Grid security solution, and so on. It has been 
deployed and used in production environment, and is one of the most widely deployed Grid
solutions in Europe.

Further development of new ARC components is pursued by the EU KnowARC project~\cite{KnowARClink}, 
based on the functionality of the current production ARC middleware. It aims at 
implementing a Web Service oriented Grid middleware which will provide higher levels of 
resource and user abstraction through a well-defined Web Service interface~\cite{KnowARCDesignlink} 
in order to provide interoperability with other Web Service oriented Grid middlewares, as well as 
other Web Service compatible applications.

The key part of the new approach is a lightweight Web 
Service container called Hosting Environment Daemon (HED), which provides hosting environment 
for various services at application level, as well as a number of modules to support flexible, 
interoperable, and efficient communication mechanism for building SOAP-based Web Services. 
The design of the HED is built around the idea of flexibility and modularity,
such that the application service developers can simply concentrate on the application 
level Web Service implementation by only using the core minimum amount of components. It 
also simplifies work on the middleware level: for example, it makes possible to implement another 
communication protocol or authentication mechanism. A system administrator can easily 
configure and deploy the middleware and application for different kinds of requirements 
without having to know much about the implementation.

The architecture of HED is illustrated in Figure~\ref{fig:HED}. In general, there 
are several components called Message Chain Component (MCC) which are in charge of implementing 
different protocol levels. For example, as shown in the message flow, the HTTP MCC 
will process a stream from the TLS MCC to parse the HTTP message and pass its body to the SOAP 
MCC, and also process the SOAP response from the SOAP MCC to generate the HTTP message 
for the TLS MCC.

Dotted line in Figure~\ref{fig:HED} shows an alternative path for the information 
to propagate among MCCs. A service administrator can configure the MCCs according to the 
interoperability requirements with a counterpart. For instance, the configuration marked 
with the dotted line is compatible with the WSE (Web Services Enhancement for .NET) SOAP 
message mechanism (c.f. WSE SoapSender and SoapReceiver)~\cite{WSElink}. Another configuration 
could be SOAP over HTTPG (HTTP over GSI) which is needed to interoperate with services 
like the Storage Resource Manager (SRM)~\cite{A.Shoshan03} service. This shows the 
flexibility of HED in terms of protocols support.

@@

HED contains a flexible security framework for implementing and enforcing
security-related functionality, such as authentication and authorization. 
Each security-related functionality can be implemented as a pluggable and
configurable component (plug-in) called \textit{SecHandler}. Each MCC or Service is usually
configured with two queues of SecHandler -- one for incoming messages and one for
outgoing messages, respectively. In Figure~\ref{fig:HED}, the ``AuthZ'' and ``AuthN'' sub-modules
inside MCCs and Services are examples of SecHandler.

%-------------------------------------------------------------------------
\section{Grid authentication using federated identity: Integrating SAML2SSO
profile}
\label{sec:intergrationSAML2SSO}
Identity federation is an emerging technology which enables the identity
information to be trustily transferred across autonomous security domains. By using identity
federation, users from one domain can access another domain without the need for a direct trust
relationship between users and accessed domains, i.e., users can use accounts from their host
domains to access external domains, and accessed domain does not need to maintain accounts
for those external users. Identity federation does not enforce specific authentication
mechanisms, therefore various types of authentication can be supported
between users and their host domains, and users are able to use their frequently used accounts or credentials to
accomplish authentication. Considering Grid use cases, we see that utilizing an existing account to
access Grid without being bothered to apply for and manipulate a X.509 credential
would be attractive for Grid users.

Shibboleth~\cite{Shiblink} is the one of the several implementations of the
identity federation, which is based on the OASIS Security Assertion Markup Language (SAML) 2.0
specifications~\cite{SAMLlink}. In terms of authentication, Shibboleth implements SAML2.0 web browser SSO
profile~\cite{SAMLprofiles}, which  defines two functional components, an Identity Provider and a Service Provider. The
Identity Provider (IdP) is responsible for creating, maintaining, and managing user identity,
while the Service Provider (SP) is responsible for controlling access to services and resources by
analysing the SAML assertions produced and issued by the IdP upon request (request from
client application to the service/resource that is protected by SP).

In the implementation of ARC middleware, SAML2.0 web browser SSO profile is
supported by implementing the functionality of service provider and web browser's user agent,
and utilizing Shibboleth IdP implementation for the functionality of identity provider. The
user agent functionality is for mimicking the behaviour of web browser's user agent for
authentication, such as the HTTP redirection and the HTTP cookie processing. Implementation
of the user agent is based on the the HTTPS client interface of ARC, since the client
interface of ARC can support different protocols which are incarnated by different MCCs.

@@

Figure~\ref{fig:SAML2SSOUML} shows how the SAML2.0 SSO profile is integrated
into the SOAP invocation between ARC WS Client and Web Service. Steps 1 to 5 in
Figure~\ref{fig:SAML2SSOUML} are similar to those steps depicted in the Web Browser SSO profile, with
the difference that we specify the way how does the service provider determine which identity
provider to use (identity provider discovery) by enabling the client to send the IdP name to the service
provider. We also specify the way how does the client (user agent)
authenticate against a service provider by implementing the form-based HTTP authentication which uses username and
password for authentication between the user agent and a service provider (rather than the X.509 client authentication),
which is compatible with the Username/Password login handler of the Shibboleth IdP.

In step 6, the service provider (SP Service) verifies and checks the SAML
response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the
AuthnStatement and AttributeStatement elements. From step 8 to step 10, the WS client first
launches the SOAP invocation via the same connection as the one which is used by the user agent to contact an SP
Service, and then the Grid/Web Service checks the AuthnStatement stored in the session context to see whether the
AuthnStatement is valid or expired. If valid, the service handles the SOAP request and returns the SOAP response to
the WS client. Note that the ``connection'' and ``session'' are interchangeably used in this paper, with
the same meaning for a TCP connection.

Note also that in the current implementation the service requires the WS client
to re-use the same TCP connection as the one used by the user agent in step 5, in order to
guarantee that the validity of the SAML2SSO result is applicable to the SOAP invocation process. On the service
side, the session re-using is accomplished by deploying the SP service and other
functional Grid/Web Service(s) together in the same service container.

There are a several benefits with integrating the SAML2.0 SSO profile for
Grid/Web service authentication. Firstly, the client certificate authentication can be switched off, so that
users would not need to apply and maintain X.509 credentials. Secondly, some implementations of
identity provider, such as Shibboleth IdP, can cache the authentication result through session management
once the user has succeeded to authenticate, and for a short period this authentication result
is valid, so that the next time the user authenticates against IdP, the user agent can just feed
IdP with the last successfully authenticated session's identity rather than feed IdP with
the username and password again. So the user (or the client on behalf of that user) can travel across multiple
security domains by only providing his name and password once, which is also a characteristic of
single sign-on. Lastly, even though there are several implementations of identity providers,
most of them are standard-compliant, so the solution implemented in ARC middleware can
interoperate with other identity provider implementations with minimum change.

%-------------------------------------------------------------------------
\section{Bridging federated identity and X.509 credential}
\label{sec:fedidtoX509}
Many widely used Grid middlewares are based on GSI which requires mutual X.509
authentication. Meanwhile, Web Service based Grid applications mostly require client
certificate authentication. To bridge the gap between federated identities and X.509 credentials, a short
lived credential service (SLCS) is implemented through which a user can get a short-lived X.509
certificate only by providing his username and password and authenticating through the SAML2SSO
profile.

Also, in order to provide single sign-on capability for services that act on
behalf of the user to invoke other services, a X.509 credential delegation service is implemented,
which is based on SOAP communication rather than the proprietary GSI communication which is used
in the delegation services such as MyProxy~\cite{myproxy}.

\subsection{Short-lived credential service}
\label{sec:slcs}
The sequence diagram of a SLCS process is the same as the diagram shown in 
Figure~\ref{fig:SAML2SSOUML}, with the SOAP request (the message of step 8 of 
Figure~\ref{fig:SAML2SSOUML}) including an X.509 request, and the SOAP response (the message of step 10 
of Figure~\ref{fig:SAML2SSOUML}) including a signed X.509 certificate and a CA certificate that is used 
to sign this X.509 certificate. In detail, the SLCS client firstly accomplishes the SAML2SSO profile, 
then sends the SOAP request to the SLCS service with the X.509 request embedded. The SLCS
service makes access control decision according to the SAML assertion stored in the session
context, and issues a certificate (with 12 hours lifetime) with SAML attributes as the X.509
certificate extension. The SLCS client then gets the SOAP response with the X.509
certificate embedded.

Since the lifetime of the short lived credential is short, it is
permissible to protect the private key by using the local file system permissions instead of encrypting
it by using a pass-phrase. Therefore, no manual interactivity is required to encrypt
the private key. Hence a single sign-on is provided, since the user only needs to enter his
password when starting a SAML2SSO profile.

A critical issue for the SLCS service is how to compose a distinguished name
(DN) for the certificate. Since the ``eduPersonPrincipalName'' (see eduPerson schema~\cite{eduSchemalink})
is identical to the user that accomplishes SAML2SSO profile, we pick up
``eduPersonPrincipalName'' from SAML attributes that returned from Shibboleth IdP, and use it to compose
the DN. A typical example of the eduPersonPrincipalName 
could be \textit{alice@example.org}. In such a case the composed DN would be,
for example,\\``/O=knowarc/OU=example.org/CN=alice''.

By using the SLCS service, a user can easily access Grid services that
require X.509 credentials on the client side from anywhere, simply by running the SLCS
client command and providing his username and password.

@@

The delegation service usage sequence diagram is shown in
Figure~\ref{fig:DelegUML}. Firstly, the client needs to delegate
the proxy certificate or the short-lived certificate to a delegation service
through steps 1 to 7.
Then this client invokes the peer Web Service by appending the delegation
information (URL of delegation service and the delegation ID corresponding to the above delegation) to the
SOAP body. The Web Service then sends a X.509 request and the delegation ID to the delegation
service which is specified by the client side, and gets back a delegated certificate
which can be composed together with the private key for a proxy certificate. The Web Service could use this
proxy certificate to represent the user to invoke another Web Service, which either could cause
one more process of delegation (steps 1 to 13) if another delegation service is specified, or
could directly start from step 8 if the same delegation service is specified as the one in the
previous delegation process.

The client only needs to delegate the certificate to the delegation service once
per session; it does not have to do delegation for subsequent SOAP invocations after the first one.

The functionalities of both client and service sides are implemented as
pluggable handlers which can be configured into the client and service configuration, so that the service or
client implementation itself does not need to be changed.

We assume that the short-lived certificate is used by the client, so that the
SLCS service together with the delegation service can provide single sign-on from user's federated
identity, while keeping the interoperability to those Grid/Web services that require mutual authentication
using certificate.

%-------------------------------------------------------------------------
\section{Performance Evaluation}
\label{sec:perfeval}
The following sections evaluate the performance of the HED versus Axis2/C, as 
well as the performance related to the SAML2SSO profile, short-lived credential
service, and delegation service.

\subsection{Performance comparison between HED and Axis2/C}
\label{sec:perhedandaxis}
Since the implementation of HED includes an alternative SOAP implementation
which is the base of the security related web services described in this paper, it is useful to
compare the performance of HED and other SOAP implementations. Although there are many SOAP
implementations such as AxisJava, gSOAP, XSOAP4, etc.., it is not the goal of this paper to do a
complete comparison between HED and all the other SOAP implementations, so we use Axis2/C (v1.5.0) to
demonstrate the performance HED.

For the service side, we implement two simple ``echoString'' services in HED and
Axis2/C. The ``echoString'' provides a simple SOAP portType which will output the same value
as it gets from input. For the client side, we developed a client by using the client API of ARC, to
invoke ``echoString'' services from Axis2/C and HED. Service invocation just sends a short
string to a service, and gets back the same string. One test involves a set of clients
which run simultaneously. Each client runs for a fixed period of time, and invokes the service as many
times as it can during the test period. Multiple clients are launched by creating multiple threads in
one test. Two values are measured: average response time and throughput. Average response
time is computed by counting the average value of the response time for each invocation.
Throughput is computed by counting number of invocations per second.

The two ``echoString'' services run in a test machine equipped with a 3GHz Intel
Pentium dual-core CPU and 2GB memory, Red Hat Enterprise Linux 5. The client runs in another
machine equipped with 2.0GHz AMD Athlon CPU and 1GB memory, Red Hat Enterprise Linux 5. The
Axis2/C service is run using the httpd module on Apache Http Server version 2.2.11 instance. We use all the
default configurations in the Apache Http Server.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{TCP_thread_all.pdf}
\caption{Performance of HED and Axis2C echoString services without TLS:
performance
changes with client numbers; session not re-used }
\label{fig:TCP_thread_all}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{TLS_thread_all.pdf}
\caption{Performance of HED and Axis2C echoString services with TLS: performance
changes with client numbers; session not re-used}
\label{fig:TLS_thread_all}
\end{figure}

Figure~\ref{fig:HED2Axis_thread10} shows results of performance comparison with
ten 
clients being launched.
Two types of experiments are shown: session re-used and session not re-used.
``Session re-used'' means that all the SOAP invocations on each client share
the same session (TCP connection); while ``Session not re-used'' means that each invocation
launches a new session.

For the ``Session re-used'' experiments, we see that the average response time
in HED is about 81ms which is twice the value for Axis2/C, and the throughput in HED is about half
of the value for Axis2/C. One observation is that the performance results of HED are quite stable, with
the average response time and throughput almost unchanged for different time points.
For the ``Session not re-used'' experiments, the average response time in HED is
almost the same as the value for Axis2/C, with the value 15ms. On the other hand, the throughput in
HED is close to the value for Axis2/C.
